---
title: "FTSA Final Project Exploration"
author: "Varshini Yanamandra"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# libraries
library(tidyverse)
library(devtools)
library(rbcpingarch)
library(ggplot2)
library(tidyr)
library(fBasics)
library(tseries)
library(astsa)
library(TSA)
library(fUnitRoots)
library(fGarch)
library(timeSeries)
library(rugarch)
library(FinTS)
library(forecast)
library(rmgarch)
library(depmixS4)
```

```{r}
#loading the data
data1 <- Y_hep
par(mfrow = c(1, 2))
hist(data1[,"Goiania"], density = TRUE, main = list("Density of Goiania Hepatitis Results"), font.main = 1)
hist(data1[,"Brasilia"], density = TRUE, main = list("Density of Brasilia Hepatitis Results"), font.main = 1)
## Both locations exhibit right skew
```

```{r}
data <- ts(data1, frequency = 12, start = 2001, end = 2018)

# plotting the data
plot(data[, 1], col = 1, ylim = c(0, 210), ylab = "Cases of Hepatitis", main = "Time Series Plots")
lines(data[, 2], col = "blue")
legend(x = "top", c("Goiania", "Brasilia"), col = c(1, "blue"), lty = c(1, 1))
```

```{r}
par(mfrow = c(2, 1))
plot(data, main = "Time Series Plot of 2 Hepatitis Locations")

TSA::acf(as.vector(data[,"Goiania"]), main = "ACF for Goiania")
pacf(as.vector(data[,"Goiania"]), main = "")
title("PACF for Goiania")
```

Clearly, both series are not stationary due to non-constant mean and variance, as seen in the time series plot. 

For the Goiania data:
The ACFs decay very slowly - this could indicate the presence of unit-root stationarity. The PACFs become mostly insignificant after lag 2, indicating that an AR component of order 2 could be used to model the series.

```{r}
par(mfrow = c(2, 1))
TSA::acf(as.vector(data[,"Brasilia"]), main = "ACF for Brasilia")
pacf(as.vector(data[,"Brasilia"]), main = "")
title("PACF for Brasilia")
```

We see a similar trend in the ACF and PACF plots for Brasilia, as well. We might try fitting an AR model with order 2 or 3 to this data.

```{r}
# testing for unit-root stationary
adfTest(as.vector(data[,"Goiania"]), lags = 12)
adfTest(as.vector(data[,"Brasilia"]), lags = 12)
```

The p-values for both series are much greater than 0.05, so we fail to reject the null hypothesis that there is the presence of unit-root stationarity in both series.

```{r}
# exploring the lagged relationship between the cases in the two cities
l = nrow(data1)
par(mfrow = c(1, 2))
plot(data1[1:l-1, 2], data1[2:l, 1], col = "red", xlab = "Brasilia(t-1)", ylab = "Goiania(t)")
plot(data1[1:l-1, 1], data1[2:l, 2], col = "blue", xlab = "Goiania(t-1)", ylab = "Brasilia(t)")
mtext("Contemporaneous Correlation in the Data", side = 3, line = -2, outer = TRUE)
```

```{r}
# LM Test
FinTS::ArchTest(data1, lag = 12)
```

Since we noticed the presence of unit-root stationarity, let us try taking a difference.

```{r}
data.diff <- diff(data, lag = 1)

plot(data.diff, main = "Time Series Plot of Differenced Data")
```

The plots looks much better in terms of the variation of the mean. We still notice that the varince is not constant, so we can not infer that the series is now stationary.

```{r}
# testing for unit-root stationary
adfTest(as.vector(data.diff[,"Goiania"]), lags = 12)
adfTest(as.vector(data.diff[,"Brasilia"]), lags = 12)
```

The p-values for both cities are now extremely small, so we reject the null hypothesis that there is the presence of unit-root stationarity in both series. The differencing has dealt with the unit-root stationarity. Let us check this by plotting the ACF plots again for both cities.

```{r}
par(mfrow = c(2, 1))
TSA::acf(as.vector(data.diff[,"Goiania"]), main = "ACF for Goiania")
TSA::acf(as.vector(data.diff[,"Brasilia"]), main = "ACF for Brasilia")
```

The ACFs now look to be well-behaved, and the presence of unit-root stationarity is now ruled out. There is a significant ACF for both cities at lag 1.

```{r}
par(mfrow = c(2, 1))
pacf(as.vector(data.diff[,"Goiania"]), main = "")
title("PACF for Goiania")

pacf(as.vector(data.diff[,"Brasilia"]), main = "")
title("PACF for Brasilia")
```

The PACFs look well-behaved, as well. Goiania has a significant PACF at lag 1, and Brasilia has significant PACFs at lags 1 and 2.

Let us find the best fit ARIMA models for the data using the 'auto.arima' function.

```{r}
fit_arima_x <- auto.arima(data[, "Goiania"], ic = "aic")
fit_arima_y <- auto.arima(data[, "Brasilia"], ic = "aic")
print("Goiania")
fit_arima_x
print("Brasilia")
fit_arima_y
```

```{r}
data1 <- as.matrix(data)
```


```{r}
# let the dccspec due to the correlated and bivariate nature of the data 
# estimate the skew and shape parameters

# trying a skewed student t with GARCH(1,1), ARIMA model set at (0,1,1) due to unit root presence:
mean.model1 <- arima(data[, "Goiania"], order = c(0, 1, 1), seasonal = list(order = c(2, 0, 0), period = 12))
mean.model2 <- arima(data[, "Brasilia"], order = c(0, 1, 1), seasonal = list(order = c(1, 0, 2), period = 12), xreg = 1:length(data[, 2]))

spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sstd")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sstd")

dccspec1 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit1 <- dccfit(dccspec1, data = data1)
dccfit1 # AIC is 16.862

aic <- 16.862
```

```{r}
##GARCH(1,0)
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 0)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sstd")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 0)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sstd")

dccspec2 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit2 <- dccfit(dccspec2, data = data1)
dccfit2 # AIC is 16.986

aic <- c(aic, 16.986)
```

```{r}
#GARCH(0,1)
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(0, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sstd")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(0, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sstd")

dccspec3 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit3 <- dccfit(dccspec3, data = data1)
dccfit3 # AIC is 17.597

aic <- c(aic, 17.597)
```

```{r}
##Trying a skewed-ged with GARCH(1,1):
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sged")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sged")

dccspec4 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit4 <- dccfit(dccspec4, data = data1)
dccfit4 # AIC is 16.946
aic <- c(aic, 16.946)
```

```{r}
#GARCH(2,0) with skewed-ged 
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(2, 0)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sged")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(2, 0)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sged")

dccspec5 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit5 <- dccfit(dccspec5, data = data1)
dccfit5 # AIC is 16.992
aic <- c(aic, 16.992)
```

```{r}
#GARCH(0,1) with skewed-ged
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(0, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sged")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(0, 1)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sged")

dccspec6 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit6 <- dccfit(dccspec6, data = data1)
dccfit6 # AIC is 26.133
aic <- c(aic, 26.133)
```

```{r}
### Trying higher order GARCH models under skewed t-dist
spec1 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(2, 2)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model1$coef), distribution.model = "sstd")
spec2 <- ugarchspec(variance.model = list(model = "sGARCH", garchOrder = c(2, 2)),
                   mean.model = list(armaOrder=c(0, 0), arfima = mean.model2$coef), distribution.model = "sstd")

dccspec7 <- dccspec(uspec = multispec(list(spec1, spec2)), dccOrder = c(1,1), distribution = "mvt")

dccfit7 <- dccfit(dccspec7, data = data1)
dccfit7 # AIC is 17.081
aic <- c(aic, 17.081)
```

```{r}
aic

### we shall disregard models
### 2-7 as they did not perform as well as model 1 or parsimony dictates
### we use simpler models with similar results.
```

```{r}
#Getting the ACF and PACF 
resid1 <- dccfit1@mfit$stdresid # gives a table with 2 columns, since the data is bivariate
# ADF test
adf.test(resid1[,1])
adf.test(resid1[,2])

plot(resid1[,1], xlab = "Time", ylab = "Value", main = "Standardized Residual Plot", col = 1, type = 'l')
lines(resid1[,2], col = "blue")
legend(x = "top", c("Goiania", "Brasilia"), col = c(1, "blue"), lty = c(1, 1))

# ACF plot
par(mfrow = c(1, 2))
TSA::acf(resid1[, 1], main = "ACF for Fit 1 - Goiania")
TSA::acf(resid1[, 2], main = "ACF for Fit 1 - Brasilia")
# PACF plot
par(mfrow = c(1, 2))
pacf(resid1[,1], main = "")
title("PACF for Fit 1 - Goiania")
pacf(resid1[,2], main = "")
title("PACF for Fit 1 - Brasilia")

resid2 <- dccfit2@mfit$stdresid
resid3 <- dccfit3@mfit$stdresid
resid4 <- dccfit4@mfit$stdresid
resid5 <- dccfit5@mfit$stdresid
resid6 <- dccfit6@mfit$stdresid
resid7 <- dccfit7@mfit$stdresid
```

```{r}
par(mfrow = c(1, 2))
TSA::acf(resid1[, 1], main = "ACF for Fit 1 - Goiania")
TSA::acf(resid1[, 2], main = "ACF for Fit 1 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid2[, 1], main = "ACF for Fit 2 - Goiania")
TSA::acf(resid2[, 2], main = "ACF for Fit 2 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid3[, 1], main = "ACF for Fit 3 - Goiania")
TSA::acf(resid3[, 2], main = "ACF for Fit 3 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid4[, 1], main = "ACF for Fit 4 - Goiania")
TSA::acf(resid4[, 2], main = "ACF for Fit 4 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid5[, 1], main = "ACF for Fit 5 - Goiania")
TSA::acf(resid5[, 2], main = "ACF for Fit 5 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid6[, 1], main = "ACF for Fit 6 - Goiania")
TSA::acf(resid6[, 2], main = "ACF for Fit 6 - Brasilia")
par(mfrow = c(1, 2))
TSA::acf(resid7[, 1], main = "ACF for Fit 7 - Goiania")
TSA::acf(resid7[, 2], main = "ACF for Fit 7 - Brasilia")
```

```{r}
par(mfrow = c(1, 2))
pacf(resid1[,1], main = "")
title("PACF for Fit 1 - Goiania")
pacf(resid1[,2], main = "")
title("PACF for Fit 1 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid2[,1], main = "")
title("PACF for Fit 2 - Goiania")
pacf(resid2[,2], main = "")
title("PACF for Fit 2 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid3[,1], main = "")
title("PACF for Fit 3 - Goiania")
pacf(resid3[,2], main = "")
title("PACF for Fit 3 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid4[,1], main = "")
title("PACF for Fit 4 - Goiania")
pacf(resid4[,2], main = "")
title("PACF for Fit 4 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid5[,1], main = "")
title("PACF for Fit 5 - Goiania")
pacf(resid5[,2], main = "")
title("PACF for Fit 5 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid6[,1], main = "")
title("PACF for Fit 6 - Goiania")
pacf(resid6[,2], main = "")
title("PACF for Fit 6 - Brasilia")

par(mfrow = c(1, 2))
pacf(resid7[,1], main = "")
title("PACF for Fit 7 - Goiania")
pacf(resid7[,2], main = "")
title("PACF for Fit 7 - Brasilia")
```

Even the best fit model out of these 7 models was not able to perform well on the data. So, we move on to try and fit the newly proposed BCP-INGARCH(1, 1) model to our data to compare the fit of this new model with the models we have already built.

MODEL USED IN THE PAPER

```{r}
# installing the package from GitHub to simulate Bivariate Conditional Poisson INGARCH(1,1) process
# devtools::install_github("luizapiancastelli/rbcpingarch")
```

```{r}
# simulating a Bivariate Conditional Poisson INGARCH(1,1) process
A = diag(c(0.3, 0.1)) #Diagonal (2x2) matrix
B = diag(c(0.2, 0.3)) #Diagonal (2x2) matrix, can be non diagonal as long as stationarity conditions satisfied
omega = c(2,1) 
phi = 0.3 
n = 500  #Time series length
Y = rBCPINGARCH(A, B, omega, phi, n)
head(Y)
cor(Y)
Y_df= data.frame('y1' =Y[,1], 'y2' = Y[,2], 't' = 1:nrow(Y))
Y_df = pivot_longer(Y_df, cols = starts_with('y'), names_to = 'serie', values_to = 'count')
ggplot(Y_df, aes(x = t, y = count, color = serie))+
  geom_line() + theme_bw() + 
  labs(y = 'Count', x = 'Time') + ylim(0, 15) + ggtitle("Simulated BCP-INGARCH(1, 1) Model Data")
```

```{r}
data("Y_hep") #Loads the Brazilian hepatites count data set
fit_diag = fit_BCP_INGARCH(Y_hep, A.diag = TRUE, B.diag = TRUE) #Diagonal model
print("DIAGONAL MODEL")
print("Maximized Parameter Values")
fit_diag$par #maximised parameter values
print("Asymptotic Standard Errors")
fit_diag$se  #asymptotic standard errors
print("Log-Likelihood")
fit_diag$loglik 
```

```{r}
fit_nd = fit_BCP_INGARCH(Y_hep, A.diag = TRUE, B.diag = FALSE) #Non-diagonal model
print("NON-DIAGONAL MODEL")
print("Maximized Parameter Values")
fit_nd$par
print("Asymptotic Standard Errors")
fit_nd$se
print("Log-Likelihood")
fit_nd$loglik
```

```{r}
print("Diagonal Model")
model_information(fit_diag)
print("Non-Diagonal Model")
model_information(fit_nd)
```

The AIC of both the diagonal and non-diagonal models is much, much lower than for any of the models we fit earlier.
